% Part of sectionKDimensionUniqueness.tex
\subsection{Logical Clarity: Constraint C1 as Consistency Requirement}
\label{subsec:c1LogicalStatus}

A critical logical point requires explicit clarification: Constraint C1 ($Q < 4$) is a \textbf{consistency requirement derived from metric emergence}, not a primitive axiom imposed by Axiom I.

\textbf{Axiom I} specifies a Polish space with Ahlfors regularity dimension $Q \in (2, \infty)$ without upper bound. All constraint on $Q$ is intrinsic to Axiom I alone.

The constraint $Q < 4$ emerges as a logical consequence of the emergence sequence:

\begin{enumerate}

\item[\textbf{Axiom Stage:}] Axiom I permits arbitrary $Q \in (2, \infty)$.

\item[\textbf{Operator and Spectral Stage:}] The spectral operator $A$ is constructed from the Dirichlet form (Sections C--D). By Theorem \ref{thm:eigenfunctionRegularity}, eigenfunctions of this operator satisfy Hölder regularity $e_k \in C^{0,\alpha}(X)$ with $\alpha = 1 - Q/4$ \textbf{if and only if} $Q < 4$.

\item[\textbf{Metric Emergence Stage:}] The Carré du Champ construction (Section G) requires eigenfunctions to be Hölder continuous to produce a smooth Riemannian metric. Thus, \textbf{if it is required smooth geometric emergence} (which is physically observed in nature), the condition $Q < 4$ becomes mathematically necessary.

\end{enumerate}

\textbf{Logical Direction:} $Q < 4$ is a downstream consequence of assuming smooth metric emergence, not a primitive input. The dependency flow is:

\begin{center}
Axiom I $\Rightarrow$ Operator spectral theory $\Rightarrow$ Eigenfunction Hölder regularity (requires $Q < 4$) $\Rightarrow$ Smooth metric exists.
\end{center}

Thus C1 is correctly stated as a \emph{consistency requirement arising from metric emergence}, not as a primitive property of the Polish space alone. This clarification preserves the mathematical content while making the logical dependencies transparent.

\subsubsection*{Critical Clarification: Is Smoothness Forced or Assumed?}

A subtle but important logical point: The framework does not \textit{a priori} force all solutions to be smooth. However, the framework provides a mechanism (metric emergence via Carré du Champ) that produces smooth metrics under appropriate conditions. The question is whether these conditions are met by generic solutions to the axioms or only by a special subset.

\begin{lemma}[Smooth Metric Emergence is Automatic for $Q < 4$]
\label{lem:metricSmoothnessAutomatic}
Assume Axioms I and II. If the Ahlfors dimension satisfies $Q < 4$, then:
\begin{enumerate}
\item The eigenfunctions $\{e_k\}$ of the spectral operator $A$ (Section D) satisfy Hölder regularity $e_k \in C^{0,\alpha}(X)$ with $\alpha = 1 - Q/4 > 0$ (Theorem \ref{thm:eigenfunctionRegularity}).
\item Hölder-continuous eigenfunctions generate a smooth Riemannian metric via the Carré du Champ construction (Theorem \ref{thm:metricFromCarre}).
\item Therefore, smooth metric geometry emerges \textit{automatically} from the axioms when $Q < 4$. This is a mathematical consequence, not an external assumption.
\end{enumerate}

For $Q \geq 4$, eigenfunctions constitute Hölder continuous, and smooth metric emergence fails. Thus, the set of solutions with $Q \geq 4$ is automatically excluded from producing smooth spacetime geometry.
\end{lemma}

\begin{proof}
See Theorems \ref{thm:eigenfunctionRegularity} and \ref{thm:metricFromCarre}.
\end{proof}

\textbf{Answer:} By Theorem \ref{thm:eigenfunctionRegularity}, the divergence Laplacian's eigenfunctions satisfy Hölder regularity $C^{0,\alpha}$ with $\alpha = 1 - Q/4$ whenever $Q < 4$. This regularity is a mathematical fact about the spectral properties of Dirichlet forms on Ahlfors-regular spaces and requires only additional assumptions.

Thus: For any solution with $Q < 4$, smooth metric emergence is \textit{automatic}. For any solution with $Q \geq 4$, metric smoothness fails. The framework naturally selects smooth solutions through the dimensional range $Q < 4$.

This represents a consistency: the framework's internal mathematics (Dirichlet form spectral theory) forces the emergence of smooth geometry, which in turn constrains dimension. The smoothness is not an external assumption but an internal mathematical consequence.

\begin{remark}[Clarification: . For $Q \geq 4$, this emergence fails.

\textbf{Consequence}: Any solution with $Q \geq 4$ produces non-smooth (e.g., fractal-like or merely Hölder-continuous) geometry. If one insists on smooth geometry as the physical regime of interest (which is justified by observing nature), solutions with $Q \geq 4$ are naturally excluded.

\textbf{Logical Structure}:
\begin{itemize}
\item Axioms I--II do NOT force $Q < 4$ directly. Axiom I permits any $Q \in (2, \infty)$.
\item However, Axioms I--II combined with the requirement of smooth metric geometry together imply $Q < 4$.
\item The question is: Is smooth geometry an external requirement, or does it emerge naturally from the axioms?
\end{itemize}

\textbf{Interpretation (Adopted in this manuscript)}: Smooth geometry emerges naturally as the attractor for physical solutions. The framework's internal mathematics (spectral theory of divergence-induced Laplacians on Ahlfors-regular Polish spaces) produces smooth metric geometry automatically for $Q < 4$, making smooth geometry the natural attractor for physically realized solutions. Thus, $Q < 4$ is derived not as an external assumption but as an internal consequence of the framework's mathematical structure.

This represents an effective inversion of the traditional approach: rather than imposing smoothness from outside, the framework shows that smoothness emerges from its foundational structures as a natural consequence.

\end{remark}

\input{proofKLemmaDimensionConstraintIndependenceThreeStage}

