% Part of sectionLMatterFieldDynamics.tex
\subsection{Nonlinear Field Equations and Solitons}
\label{subsec:nonlinearFieldEquationsAndSolitons}

\begin{definition}[Nonlinear Matter Hamiltonian]
\label{def:nonlinearHamiltonian}
The matter field Hamiltonian with self-interaction:
\begin{equation}
H_{\text{matter}}[\psi] := \int_X \left[\frac{1}{2}\Gamma(\psi, \psi) + U(|\psi|^2)\right] d\mu,
\end{equation}
where $\Gamma(\psi, \psi)$ is the Carré du Champ kinetic term (Definition \ref{def:carreDuChamp}) and $U(|\psi|^2) := \alpha |\psi|^4 + \beta |\psi|^6$ with $\alpha, \beta > 0$ is the self-interaction potential.
\end{definition}

\begin{theorem}[Nonlinear Wave Equation in Lorentzian Signature]
\label{thm:nonlinearWaveEquation}
Varying the action $S = \int dt \int_X [\frac{1}{2}(\partial_t \psi)^2 - H_{\text{matter}}[\psi]] d\mu$ yields:
\begin{equation}
\Box \psi + 2U'(|\psi|^2) \psi = 0,
\end{equation}
where $\Box := -\partial_t^2 + \Delta_g$ is the d'Alembertian with respect to the Lorentzian metric $g$ from Theorem \ref{thm:lorentzianSignature}.

For static configurations ($\partial_t \psi = 0$):
\begin{equation}
-\Delta_g \psi + 2U'(|\psi|^2) \psi = \mu \psi,
\end{equation}
where $\mu$ is a Lagrange multiplier (chemical potential).

\begin{proof}
\input{proofLTheoremNonlinearWaveEquation}
\end{proof}
\end{theorem}

\begin{theorem}[Solitonic Particle Solutions Exist]
\label{thm:solitonicParticles}
For $\alpha = 1, \beta = 1$ (or more generally any $\alpha, \beta > 0$), the static nonlinear equation admits solitonic solutions:
\begin{equation}
\psi_{\text{soliton}}(x) = \phi_0(x) e^{i\theta},
\end{equation}
where $\phi_0: X \to \mathbb{R}^+$ is exponentially localized:
\begin{equation}
\phi_0(x) \sim e^{-\lambda_0 d_g(x, x_0)} \quad \text{as } d_g(x, x_0) \to \infty
\end{equation}
for a center $x_0 \in X$ and decay rate $\lambda_0 > 0$.

These solutions possess:

\begin{enumerate}
\item \textbf{Finite Energy.} $H_{\text{matter}}[\psi_{\text{soliton}}] < \infty$.

\item \textbf{Stability.} They are local minima of $H_{\text{matter}}$ subject to fixed charge $Q := \int |\psi|^2 d\mu$.

\item \textbf{Topological Protection.} The winding number $\nu := \frac{1}{2\pi} \oint_{\partial D} d\theta \in \mathbb{Z}$ is conserved.

\item \textbf{Rest Mass.} $M_{\text{rest}} := H_{\text{matter}}[\psi_{\text{soliton}}]$ defines the particle rest mass.

\item \textbf{Relativistic Dispersion.} Time-dependent solutions $\psi(x,t) = \phi_0(x - vt) e^{iEt}$ exhibit relativistic dispersion relations.
\end{enumerate}

\begin{proof}
\input{proofLTheoremSolitonicParticles}
\end{proof}
\end{theorem}

