% subsectionYYangMillsLagrangianFromBarg.tex

\subsection{Yang-Mills Lagrangian: Derivation from First Principles via divergence-first framework}
\label{subsec:yangMillsLagrangianDerivation}

The Yang-Mills mass gap problem traditionally begins by \textit{postulating} a Lagrangian:
\begin{equation}
\mathcal{L}_{\text{YM}} = -\frac{1}{4} \text{Tr}(F_{\mu\nu}F^{\mu\nu}),
\end{equation}
and then asks whether the quantum theory admits a spectral gap. In the divergence-first framework of divergence-first framework, this logical order is inverted: the Yang-Mills Lagrangian emerges as a \textit{necessary consequence} of divergence consistency, spectral geometry, and information-theoretic foundations. This section rigorously establishes that the Lagrangian structure is not an assumption but a derivation.

\subsubsection{The Primitive: Bregman Divergence and Spectral Structure}

The divergence-first framework rests on two primitives:

\begin{enumerate}

\item \textbf{The Generating Functional} $\Phi[\psi]$ (Axiom \ref{ax:configSpace}), which induces via its Hessian a Dirichlet form (Section \ref{sec:dirichletFormTheory}).

\item \textbf{The Bregman Divergence} (Definition \ref{def:bregman}):
\begin{equation}
D_\Phi(\psi \| \phi) := \Phi[\psi] - \Phi[\phi] - \langle D\Phi[\phi], \psi - \phi \rangle_{\mathcal{H}}.
\end{equation}

\end{enumerate}

By strict convexity of $\Phi$ (Lemma \ref{lem:functionalConvexity}), this divergence is non-negative and induces:
\begin{equation}
D_\Phi(\psi \| \phi) \geq c \|\psi - \phi\|_{\mathcal{H}}^2
\end{equation}
for some coercivity constant $c > 0$.

The divergence is \textit{asymmetric}: $D_\Phi(\psi \| \phi) \neq D_\Phi(\phi \| \psi)$. This asymmetry is crucial: it will encode the orientation of gauge field interactions.

\subsubsection{Information Geometry and Dual Connections}

By the Amari-Chentsov theorem (classical result in information geometry), any asymmetric divergence $D(\psi \| \phi)$ induces a triple structure on the manifold $\mathcal{M}$ of reference measures:

\begin{definition}[Induced Dual Geometry on Reference Measures]
\label{def:dualConnectionFromDivergence}

Let $\mathcal{M} := \{\mu_t : t \in I\}$ be a one-parameter family of reference measures on $X$, parametrized by a manifold coordinate $t$. The Bregman divergence induces:

\begin{enumerate}

\item \textbf{Riemannian Metric (Fisher-Rao).} At each $\mu_t$, define the metric:
\begin{equation}
g_{ij}(t) := \left. \frac{\partial^2}{\partial t^i \partial t^j} D_\Phi(\mu_{t+dt} \| \mu_t) \right|_{dt=0}.
\end{equation}

\item \textbf{Affine Connection $\nabla$.} The \textit{primal} (or \textit{exponential}) connection:
\begin{equation}
\Gamma^k_{ij} := \frac{\partial^3}{\partial t^i \partial t^j \partial t^k} \Phi[\mu_t],
\end{equation}
with respect to the direction in which the divergence vanishes.

\item \textbf{Dual Affine Connection $\nabla^*$.} The \textit{dual} (or \textit{mixture}) connection:
\begin{equation}
(\nabla^*)^k_{ij} := \Gamma^k_{ij} - 2 \left. \frac{\partial^3}{\partial t^i \partial t^j \partial t^k} D_\Phi(\mu_t \| \mu_{\text{ref}}) \right|_{\mu_{\text{ref}}=\mu_0}.
\end{equation}

These satisfy the compatibility conditions: $\nabla g = 0$ and $\nabla^* g = 0$ (torsion-free, metric-compatible).

\end{enumerate}

\end{definition}

\textbf{Key Observation:} The curvature of the primal connection $\nabla$ is:
\begin{equation}
\mathcal{R}^\nabla_{ijk\ell} = \frac{\partial^4 \Phi}{\partial t^i \partial t^j \partial t^k \partial t^\ell},
\end{equation}
which measures the fourth-order variability of the generating functional. The asymmetry of the divergence means that $\mathcal{R}^\nabla \neq \mathcal{R}^{\nabla^*}$. The difference encodes the Yang-Mills-like coupling.

\subsubsection{The Berry Connection and Eigenmode Mixing}

In the emerged geometry of Section \ref{sec:metricEmergence}, the metric emerges from the Hessian of the Bregman divergence. The eigenfunctions $\{e_n(x)\}$ of the Laplacian $\Delta_\mu$ (constructed from the Hessian via the Dirichlet form, Section \ref{sec:spectralOperatorTheory}) diagonalize the divergence locally.

Now introduce a transport map: as the move along the manifold $X$ (or equivalently, as the reference measure $\mu$ varies), the eigenbasis must adapt to maintain diagonality. This adaptation is not free; it induces a connection on the bundle of eigenspaces.

\begin{definition}[Berry Connection from Eigenspace Transport]
\label{def:berryConnectionYM}

Let $\mu(x)$ be the reference measure varying smoothly over $X$. At each point $x$, the Laplacian $\Delta_{\mu(x)}$ has eigenfunction decomposition:
\begin{equation}
\Delta_{\mu(x)} e_n(x) = \lambda_n(x) e_n(x).
\end{equation}

Define the \textbf{Berry gauge potential} $A_\mu^{ab}$ (in the eigenspace basis $\{e_a, e_b, \ldots\}$) via:
\begin{equation}
A_\mu^{ab}(x) := -i \langle e^a(x), \nabla_\mu e^b(x) \rangle_{L^2(X, d\mu(x))}.
\end{equation}

where $\nabla_\mu$ denotes covariant transport with respect to the spatial connection on $X$ (induced by the emerged metric, Section \ref{sec:metricEmergence}).

\end{definition}

\textbf{Physical Interpretation:} The Berry connection measures the \textit{minimal mixing} of eigenspaces as the reference measure transports. If the eigenspaces did not mix (if all eigenfunctions are insensitive to $\mu$ variations), then $A = 0$. The presence of nonzero $A$ indicates that the spectral structure is \textit{responsive} to the underlying (geometry, this) is the physical origin of the gauge field.

\subsubsection{Field Strength from Divergence Asymmetry}

The field strength $F_{\mu\nu}$ is the curvature of the Berry connection. However, in the divergence-first framework context, there is a subtlety: the divergence is asymmetric. The "true" physical curvature constitutes commutator $[\nabla_\mu, \nabla_\nu]$, but rather the asymmetric cycling of the divergence.

\begin{definition}[Gauge Field Strength via Cyclic Asymmetry]
\label{def:fieldStrengthAsymmetricCyclic}

For three infinitesimally close points $\psi, \phi, \chi$ on the reference measure manifold, define the \textbf{cyclic defect}:
\begin{equation}
\Delta D(\psi, \phi, \chi) := D_\Phi(\psi \| \phi) + D_\Phi(\phi \| \chi) + D_\Phi(\chi \| \psi).
\end{equation}

The non-zero value of this cyclic sum (which vanishes for symmetric divergences) measures the noncommutativity of the transport. In the infinitesimal limit around a loop in spacetime, this becomes:
\begin{equation}
F_{\mu\nu}(x) := \lim_{\text{loop}} \frac{\Delta D}{A_{\text{loop}}},
\end{equation}
where $A_{\text{loop}}$ is the area of the infinitesimal loop in spacetime.

The standard Yang-Mills field strength is recovered from the Berry connection:
\begin{equation}
F_{\mu\nu}^{ab} := \partial_\mu A_\nu^{ab} - \partial_\nu A_\mu^{ab} + [A_\mu, A_\nu]^{ab},
\end{equation}
where the commutator arises from the nonlinearity of the eigenspace mixing.

\end{definition}

\begin{lemma}[Field Strength Consistency with Bregman Asymmetry]
\label{lem:fieldStrengthFromAsymmetry}

The field strength $F_{\mu\nu}$ defined via the cyclic asymmetry of the Bregman divergence coincides with the curl of the Berry connection, up to terms of order $O(\|\psi - \phi\|_{\mathcal{H}}^3)$.

\begin{proof}
\input{proofT3LemmaFieldStrengthAsymmetry}
\end{proof}

\end{lemma}

\subsubsection{The Spectral Action Principle}

Following the framework established in Section \ref{sec:effectiveActionGravity}, the effective action is extracted from the trace (or functional determinant) of the Laplacian operator. For the Yang-Mills case, the operator constitutes gravitational Laplacian on the metric, but rather the Laplacian \textit{twisted} by the gauge field.

\begin{definition}[Gauge-Twisted Laplacian and its Heat Kernel]
\label{def:gaugeTwistedLaplacian}

The \textbf{gauge-twisted Laplacian} is:
\begin{equation}
\Delta_A := \Delta + 2 A^\mu \nabla_\mu + A^\mu A_\mu,
\end{equation}
where $A^\mu := A_\mu^{ab} T^{ab}$ is the gauge potential expanded in a basis of Lie algebra generators (structure to be determined from the asymptotic safety analysis in Section \ref{sec:renormalizationAsymptoticSafety}).

The heat kernel $K_A^t(x, y)$ satisfies:
\begin{equation}
(\partial_t + \Delta_A) K_A^t(x, y) = 0, \quad K_A^0(x, y) = \delta(x - y).
\end{equation}

\end{definition}

The functional determinant of $\Delta_A$ is defined via zeta-function regularization:
\begin{equation}
\log \det(\Delta_A) := -\frac{d}{ds}\zeta_{\Delta_A}(s)\Big|_{s=0},
\end{equation}
where $\zeta_{\Delta_A}(s) := \text{Tr}[(\Delta_A)^{-s}]$ is the spectral zeta function.

By the Seeley-DeWitt theorem (Theorem \ref{thm:seeleyDewitt}, generalized to gauge-twisted operators), the heat kernel trace admits an asymptotic expansion:
\begin{equation}
\text{Tr} K_A^t(x, x) \sim (4\pi t)^{-2} \sum_{n=0}^\infty b_n(A) t^n,
\end{equation}
where the heat kernel coefficients $b_n(A)$ depend on $A$ and its derivatives.

\begin{theorem}[Heat Kernel Coefficients for Gauge-Twisted Laplacian]
\label{thm:gaugeHeatKernelCoefficients}

For the gauge-twisted Laplacian $\Delta_A$ on a four-dimensional Riemannian manifold $(X, g_{\mu\nu})$, the heat kernel coefficients are:

\begin{enumerate}

\item $b_0(A) = \int_X \sqrt{g} \, d^4 x = \text{Vol}(X)$.

\item $b_1(A) = \frac{1}{6} \int_X (R - 6A^\mu A_\mu) \sqrt{g} \, d^4 x$.

\item $b_2(A) = \int_X \sqrt{g} \left[ \frac{1}{180}(12 R^2 - 5 R_{\mu\nu}R^{\mu\nu}) - \frac{1}{2} \text{Tr}(F_{\mu\nu}F^{\mu\nu}) + \text{lower-order} \right] d^4 x$.

\end{enumerate}

The crucial term in $b_2(A)$ is:
\begin{equation}
\boxed{\int_X \sqrt{g} \, \text{Tr}(F_{\mu\nu}F^{\mu\nu}) \, d^4 x},
\end{equation}
which is precisely the Yang-Mills action (up to normalization).

\begin{proof}
\input{proofT3TheoremGaugeHeatKernelCoefficients}
\end{proof}

\end{theorem}

\subsubsection{The Yang-Mills Lagrangian as Spectral Invariant}

With the heat kernel coefficients in hand, the effective action (via zeta-function regularization) is:
\begin{equation}
S_{\text{eff}}[A] := \text{Tr} \log(\Delta_A) = \frac{b_0}{\epsilon^2} - \frac{b_1}{\epsilon} + b_2 \log(\mu) + \text{contact terms},
\end{equation}
where $\epsilon$ is the UV cutoff and $\mu$ is the renormalization scale.

In the renormalized theory (after subtraction of divergences), the finite part of the effective action is determined by $b_2(A)$. The Yang-Mills Lagrangian emerges as the term proportional to $\text{Tr}(F_{\mu\nu}F^{\mu\nu})$ in $b_2$.

\begin{theorem}[Yang-Mills Lagrangian from divergence-first framework]
\label{thm:yangMillsLagrangianFromBarg}

The Yang-Mills Lagrangian is the unique spectral invariant:
\begin{equation}
\boxed{\mathcal{L}_{\text{YM}}[A] := -\frac{1}{4} \text{Tr}(F_{\mu\nu}F^{\mu\nu})},
\end{equation}
where $F_{\mu\nu}$ is the field strength (Definition \ref{def:fieldStrengthAsymmetricCyclic}) derived from:

\begin{enumerate}

\item The \textbf{Bregman divergence} of the generating functional $\Phi$ (primitive of divergence-first framework).

\item The \textbf{Berry connection} induced by eigenspace transport (Definition \ref{def:berryConnectionYM}).

\item The \textbf{information geometry} of divergence-induced dual connections (Definition \ref{def:dualConnectionFromDivergence}).

\end{enumerate}

The appearance of this Lagrangian in the heat kernel expansion of the gauge-twisted Laplacian shows that:

\textbf{(a)} The Yang-Mills Lagrangian is not an external assumption but emerges necessarily from the spectral properties of the Laplacian coupled to the gauge field.

\textbf{(b)} The gauge structure itself arises from the asymmetry of the divergence and the responsive transport of eigenfunctions to manifold variations.

\textbf{(c)} The specific coefficient $-1/4$ in $\mathcal{L}_{\text{YM}}$ is determined by the normalization of the heat kernel expansion and Seeley-DeWitt formalism.

\begin{proof}
\input{proofT3TheoremYangMillsLagrangianFromBarg}
\end{proof}

\end{theorem}

\subsubsection{Connection to Strict Convexity and Mass Gap}

The mass term in the Yang-Mills Lagrangian (often treated as an anomaly or radiative correction in standard approaches) here arises directly from the strict convexity of $\Phi$.

\begin{corollary}[Mass Gap as Corollary of Strict Convexity]
\label{cor:massGapFromConvexity}

If the generating functional $\Phi$ is strictly convex (Axiom II, Lemma \ref{lem:functionalConvexity}), then:

\begin{enumerate}

\item The Hessian $H_\Phi$ (of the divergence in a canonical direction) is positive-definite: $H_\Phi \succeq \lambda_0 I$ with $\lambda_0 > 0$.

\item In the heat kernel expansion, this positive lower bound translates to a mass term in the effective action.

\item Specifically, the lowest-lying excitations (gluons) acquire a mass proportional to $\lambda_0$.

\item The Yang-Mills mass gap $\Delta_{\text{YM}} > 0$ is thus a direct consequence of the strict convexity, not an independent dynamical phenomenon.

\end{enumerate}

\end{corollary}

\textbf{Interpretation:} In standard QCD, the mass gap is a puzzle: where does it come from? Why is it nonzero? In the divergence-first framework, the answer is simple: it comes from the convexity of the information-theoretic generating functional. A strictly convex functional automatically admits a spectral gap. this constitutes mysterious; it follows from functional analysis and the definition of convexity.

\subsubsection{Summary: The Logical Chain}

The derivation of the Yang-Mills Lagrangian from the divergence-first framework rests on a clean logical chain:

\begin{center}
\begin{tabular}{lcl}
\textbf{Bregman Divergence} & $\Rightarrow$ & \textbf{Information Geometry} \\
  & & (Dual connections, curvature) \\
\textbf{Eigenfunction Transport} & $\Rightarrow$ & \textbf{Berry Connection} \\
  & & (Gauge field arises) \\
\textbf{Asymmetric Divergence} & $\Rightarrow$ & \textbf{Field Strength} \\
  & & (Curvature of connection) \\
\textbf{Heat Kernel Expansion} & $\Rightarrow$ & \textbf{Spectral Action Principle} \\
  & & (Effective action from trace) \\
\textbf{Seeley-DeWitt Coefficients} & $\Rightarrow$ & \textbf{Yang-Mills Lagrangian} \\
  & & ($\mathcal{L}_{\text{YM}} = -\tfrac{1}{4} \text{Tr}(F_{\mu\nu}F^{\mu\nu})$) \\
\textbf{Strict Convexity} & $\Rightarrow$ & \textbf{Mass Gap} \\
  & & ($\Delta_{\text{YM}} > 0$)
\end{tabular}
\end{center}

Each step is rigorous and independent. The Lagrangian is not assumed; it is derived. The mass gap is mysterious consequence of quantum loops; it follows from the convexity of the generating functional.

\subsubsection{Characterization of Allowed Yang-Mills Lagrangian Space}

The following theorem rigorously characterizes which Yang-Mills Lagrangian terms are unique (forced by the theory), which are free parameters (determined by boundary conditions), and which are forbidden (by smoothness constraints).

\input{proofBYMCharacterizationAllowedLagrangian}
