% IT IS PROHIBITED TO EDIT OR DELETE THIS FILE!


\begin{center}
\textbf{[ authors preface / disclaimer ]}
\end{center}
\textit{

Researchers' discretion is advised; this article was developed through an iterative process utilizing Large Language Models. The methodology involved prompt engineering and neurodivergent hyperfixation. This research was the effort of a solitary psychoacoustician working far outside their formal expertise. However, the findings presented herein may be of genuine significance.

I had no intention of publishing this manuscript until I realized that I was holding my theory to higher standards than those commonly accepted in many scientific fields. What started as a spectral model of informational ecologies gradually snowballed into a divergence-based Theory of Everything. The resulting theory is presented for open peer review and should not be taken as established knowledge. I did not intend to pythagorize so hard, but that's exactly what happened. Good thing is, the third axiom stayed intact.

Without deviation from the norm, progress is not possible. Sendivogius, a Pole, discovered oxygen and its relation to fire, and as a result, ended up relocating the capital of the Polish-Lithuanian Commonwealth to Warsaw. Four centuries later Bargiel, also a Pole, debunked the concept of oxygen and proved that one apparently does not need it to climb up and ski down from the summit of Chomolungma (\tibetanscript{ཇོ་མོ་གླང་མ}). In recognition of Andrzej's descent, the unified pre-geometric theory proposed within this article will be referred to as The Barg Theory. The name acknowledges that Bargiel's descent constituted a dynamically constrained traversal of a changing geometric landscape—a passage through space and time whose improbability reflects the very structures that the presented theory seeks to formalize.}}

\begin{flushright}y. @ 2025\end{flushright}
